Modelling potential environmental impacts of science activity in Antarctica

### 2024-12-10 paper/paper.qmd
fixed minor typo
latest edits to paper draft

```{r}
library(ggplot2)
library(ggpubr)
library(grid)
library(cols4all)
library(tidyr)
library(dplyr)
library(igraph)
library(sf)
library(terra)
library(akima)
library(stringr)
library(ggnewscale)
# library(whitebox)
library(ggspatial)
```

@@

GPS data were processed (see @fig-gps-processing) to make them better suited for use in the estimation of hiking functions.

@@

# geologies data
geologies <- 
  st_read(str_glue("{data_folder}/ata-scar-geomap-geology-v2022-08-clipped.gpkg")) |> 
  dplyr::select(POLYGTYPE) |>
  st_filter(bb |> st_as_sfc()) |> 
  mutate(cover = factor(POLYGTYPE))

@@

GPS data and a crudely estimated hiking functions before and after filtering of the data.
:::

## Methods and results

### Hiking functions {#sec-hiking-function-estimation}
Fit three alternative functional forms: exponential Tobler (1993, pp 1–4, based on data from Imhof 1950), Gaussian Irmischer and Clarke (2018), and Student’s t distribution, the last of these as a heavier-tailed alternative to Gaussian.
Other approaches rely on rule-of-thumb hiking time calculations such as...

The Gaussian form offered the best overall fits to our data. 
Although Tobler’s exponential form is much used it is somewhat implausible that hiking speeds would peak sharply at a very specific slope angle, and so we adopt the Gaussian form.

\[FIGURE showing hiking function fits about here\]

In previous work researchers have applied a ground cover penalty cost to a base hiking function to estimate traversal times. 
In our case there are effectively only two terrain types, moraines (gravel) and bare rock. 
This makes the option of fitting different functions to each attractive and reasonable, and the results of fitting distinct functions to each support this approach. 
The peak speed on bare rock is attained on steeper downhill slopes than on gravel, perhaps indicative of the greater care required on downhill gravel slopes. 
Meanwhile the highest speeds on level terrain are attained on gravel. 
Our fitted functions are shown below.

\[FIGURE comparing moraine and rock hiking functions about here\]

>>>>

```{r}
library(ggplot2)
library(ggpubr)
library(grid)
library(cols4all)
library(tidyr)
library(dplyr)
library(igraph)
library(sf)
library(terra)
library(akima)
library(stringr)
library(ggnewscale)
# library(whitebox)
library(ggspatial)
library(minpack.lm)
```

@@

GPS data were processed to make them better suited for use in the estimation of hiking functions.

@@

# geologies data
geologies <- 
  st_read(str_glue("{data_folder}/ata-scar-geomap-geology-v2022-08-clipped.gpkg")) |> 
  dplyr::select(POLYGTYPE) |>
  st_filter(bb |> st_as_sfc()) |> 
  mutate(cover = factor(str_to_title(POLYGTYPE)))

@@

GPS data and crudely estimated hiking functions before and after filtering the to remove fixes associated iwht non-purposive movement.
:::

## Methods and results
### Hiking functions {#sec-hiking-function-estimation}
We fit three alternative functional forms to the cleaned GPS data: exponential [@Tobler1993], Gaussian [following @Irmischer2018], and Lorentz [following @Campbell2019] using the Levenburg-Marquardt algorithm [@More1978] as provided by the `nlsLM` function in the `minpack.lm` R package [@Elzhov2022].
The raw data and fitted curves are shown in @fig-comparison-of-hiking-functions.

```{r}
get_gaussian_hiking_function <- function(df) {
  nlsLM(speed_km_h ~ a * dnorm(slope_h, m, s), data = df,
        start = c(a = 5, m = 0, s = 0.5))
}
get_tobler_hiking_function <- function(df) {
  nlsLM(speed_km_h ~ a * exp(-b * abs(slope_h + c)), data = df,
        start = c(a = 5, b = 3, c = 0.05))  
}
get_lorentz_hiking_function <- function(df) {
  nlsLM(speed_km_h ~ a /(pi * b * (1 + ((slope_h - c) / b) ^ 2)), 
        data = df, start = c(a = 5, b = 1, c = -0.05))
}

# makes a prediction data frame with x, y values
# slopes is a DF with one column called slope_h
get_model_prediction_df <- function(m, slopes) {
  data.frame(x = slopes, y = predict(m, data.frame(slope_h = slopes$slope_h)))
}

slopes <- data.frame(slope_h = -150:150 / 100)

hiking_functions <- list(
  Exponential = get_tobler_hiking_function,
  Gaussian = get_gaussian_hiking_function,
  Lorentz = get_lorentz_hiking_function
)
covers <- c("Moraine", "Rock")

models <- list()
predictions <- list()
inputs <- list()
i <- 1
for (name in names(hiking_functions)) {
  for (cover_type in covers) {
    df <- gps_geol_purposive |> 
      filter(cover == cover_type) 
    m <- hiking_functions[[name]](df)
    models[[cover_type]][[name]] <- m
    predictions[[i]] <- get_model_prediction_df(m, slopes) |>
      mutate(cover = cover_type, Model = name)
    inputs[[i]] <- df |>
      mutate(cover = cover_type, Model = name)
    i <- i + 1
  }
}
model_predictions <- bind_rows(predictions)
input_data <- bind_rows(inputs)
```

```{r}
#| label: fig-comparison-of-hiking-functions
#| fig-cap: Three possible hiking functions applied to GPS data split by land cover.
ggplot() +
  geom_boxplot(data = gps_geol_purposive |> filter(cover != "Ice"),
               aes(x = slope_h_round, y = speed_km_h, group = slope_h_round),
               outlier.size = 0.35, colour = "grey", linewidth = 0.5) +
  # geom_point(data = input_data, aes(x = slope_h, y = speed_km_h)) + 
  geom_line(data = model_predictions, aes(x = slope_h, y = y, colour = Model)) +
  xlab("Slope, rise over run") + ylab("Speed, km/h") +
  facet_wrap( ~ cover, ncol = 3) +
  theme_minimal()
```

The Lorentz function offers a marginal improvement in the model fit in comparison with the Gaussian function, while both are clearly better than the exponential form. 
However, the improvement offered by the Lorentz function over the Gaussian is marginal: residual standard error 1.489 vs 1.491 on Moraine, and 1.487 vs 1.488 on Rock, and inspection of the curves shows that estimated hiking speeds for the Gaussian functions are much closer to a plausible zero on very steep slopes. 
We therefore chose to adopt Gaussian hiking functions for the remainder of the present work.

In previous work researchers have applied a ground cover penalty cost to a base hiking function to estimate traversal times.
We instead, as shown, estimate different hiking functions for the two ground cover types present.
The peak speed on rock is attained on steeper downhill slopes than on moraines, perhaps indicative of the greater care required on downhill gravel slopes. 
Meanwhile the highest speeds on level terrain are attained on moraines.

```{r}
# make a all data function
df <- gps_geol_purposive 
m3 <- get_gaussian_hiking_function(df)
models[["All"]][["Gaussian"]] <- m
model_predictions <- model_predictions |>
  bind_rows(get_model_prediction_df(m, slopes) |> 
              mutate(cover = "All", model = "Gaussian"))
input_data <- input_data |>
  bind_rows(df |> mutate(cover = "All", model = "Gaussian"))

```

```{r}
#| output: false

# see https://rmazing.wordpress.com/2013/08/14/predictnls-part-1-monte-carlo-simulation-confidence-intervals-for-nls-models/
predictNLS <- function(object, newdata, level = 0.95, nsim = 10000, ...) {
  require(MASS, quietly = TRUE)
  ## get right-hand side of formula
  RHS <- as.list(object$call$formula)[[3]]
  EXPR <- as.expression(RHS)
  ## all variables in model
  VARS <- all.vars(EXPR)
  ## coefficients
  COEF <- coef(object)
  ## extract predictor variable    
  predNAME <- setdiff(VARS, names(COEF))  
  ## take fitted values, if 'newdata' is missing
  if (missing(newdata)) {
    newdata <- eval(object$data)[predNAME]
    colnames(newdata) <- predNAME
  }
  ## check that 'newdata' has same name as predVAR
  if (names(newdata)[1] != predNAME) stop("newdata should have name '", predNAME, "'!")
  ## get parameter coefficients
  COEF <- coef(object)
  ## get variance-covariance matrix
  VCOV <- vcov(object)
  ## augment variance-covariance matrix for 'mvrnorm' 
  ## by adding a column/row for 'error in x'
  NCOL <- ncol(VCOV)
  ADD1 <- c(rep(0, NCOL))
  ADD1 <- matrix(ADD1, ncol = 1)
  colnames(ADD1) <- predNAME
  VCOV <- cbind(VCOV, ADD1)
  ADD2 <- c(rep(0, NCOL + 1))
  ADD2 <- matrix(ADD2, nrow = 1)
  rownames(ADD2) <- predNAME
  VCOV <- rbind(VCOV, ADD2) 
  ## iterate over all entries in 'newdata' as in usual 'predict.' functions
  NR <- nrow(newdata)
  respVEC <- numeric(NR)
  seVEC <- numeric(NR)
  varPLACE <- ncol(VCOV)   
  ## define counter function
  counter <- function (i) {
    if (i%%10 == 0) 
      cat(i)
    else cat(".")
    if (i%%50 == 0) 
      cat("\n")
    flush.console()
  }
  outMAT <- NULL 
  for (i in 1:NR) {
    counter(i)
    ## get predictor values and optional errors
    predVAL <- newdata[i, 1]
    if (ncol(newdata) == 2) predERROR <- newdata[i, 2] else predERROR <- 0
    names(predVAL) <- predNAME  
    names(predERROR) <- predNAME  
    ## create mean vector for 'mvrnorm'
    MU <- c(COEF, predVAL)
    ## create variance-covariance matrix for 'mvrnorm'
    ## by putting error^2 in lower-right position of VCOV
    newVCOV <- VCOV
    newVCOV[varPLACE, varPLACE] <- predERROR^2
    ## create MC simulation matrix
    simMAT <- mvrnorm(n = nsim, mu = MU, Sigma = newVCOV, empirical = TRUE)
    ## evaluate expression on rows of simMAT
    EVAL <- try(eval(EXPR, envir = as.data.frame(simMAT)), silent = TRUE)
    if (inherits(EVAL, "try-error")) stop("There was an error evaluating the simulations!")
    ## collect statistics
    PRED <- data.frame(predVAL)
    colnames(PRED) <- predNAME   
    FITTED <- predict(object, newdata = data.frame(PRED))
    MEAN.sim <- mean(EVAL, na.rm = TRUE)
    SD.sim <- sd(EVAL, na.rm = TRUE)
    MEDIAN.sim <- median(EVAL, na.rm = TRUE)
    MAD.sim <- mad(EVAL, na.rm = TRUE)
    QUANT <- quantile(EVAL, c((1 - level)/2, level + (1 - level)/2))
    RES <- c(FITTED, MEAN.sim, SD.sim, MEDIAN.sim, MAD.sim, QUANT[1], QUANT[2])
    outMAT <- rbind(outMAT, RES)
  }
  colnames(outMAT) <- c("fit", "mean", "sd", "median", "mad", names(QUANT[1]), names(QUANT[2]))
  rownames(outMAT) <- NULL
  cat("\n")
  return(outMAT)  
}

conf <- 0.95
pred1 <- predictNLS(models[["Moraine"]][["Gaussian"]], newdata = slopes, level = conf) |>
  as.data.frame() |>
  mutate(x = slopes$slope_h, cover = "Moraine")
pred2 <- predictNLS(models[["Rock"]][["Gaussian"]], newdata = slopes, level = conf) |>
  as.data.frame() |>
  mutate(x = slopes$slope_h, cover = "Rock")
pred3 <- predictNLS(m3, newdata = slopes, level = conf) |>
  as.data.frame() |>
  mutate(x = slopes$slope_h, cover = "All")
predictions <- bind_rows(pred1, pred2, pred3)
```

Overplotting of the hiking functions including an additional model fitted to all the data, confirms that the fitted functions are sufficiently different to retain separate models for each ground cover (see @fig-models-by-cover-compared).
Plotting both functions in the same graph makes clearer the difference in maximum speed and slope at maximum speed associated with each ground cover.

```{r}
#| label: fig-models-by-cover-compared
#| fig-cap: The hiking functions for All, Moraine and Rock ground covers compared, including 95% confidence intervals derived by Monte-Carlo simulation.
ggplot() +
  geom_ribbon(data = predictions |> filter(cover != "All"), 
              aes(x = x, ymin = `2.5%`, ymax = `97.5%`, group = cover, fill = cover), 
              alpha = 0.35, linewidth = 0) + 
  scale_fill_brewer(palette = "Set1", name = "Cover") +
  geom_ribbon(data = predictions |> filter(cover == "All"), 
              aes(x = x, ymin = `2.5%`, ymax = `97.5%`), 
              colour = "black", lty = "dashed", fill = "#00000000", linewidth = 0.35) +
  geom_line(data = predictions |> filter(cover == "All"), aes(x = x, y = mean, colour = cover)) +
  scale_colour_manual(name = "", breaks = c("All"), values = c("black")) +
  xlab("Slope, rise over run") + ylab("Speed, km/h") +
  theme_minimal()
```
